---
title: "ps5-markdown-plot"
author: "selina"
date: "2/13/2023"
output: html_document
---

1. Load and check data
```{r}
getwd()
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```
(2). The _gapminder_ data contained `r nrow(gapminder)` rows and `r ncol(gapminder)` columns. 

(3):
```{r}
gapminder %>% 
  sample_n(5)
```


2. Descriptive statistics

(1). For 3-letter country code, there are `r length(unique(gapminder$iso3))` countries; for 2-letter country code, there are `r length(unique(gapminder$iso2))` countries; and for names, there are `r length(unique(gapminder$name))` countries. 

(2). (a)
```{r}
gapminder %>%
  group_by(name, iso2) %>% 
  summarise(n_distinct(iso2), n()) %>%
  arrange(desc(n()))
```

(b)
```{r}
gapminder %>%
  group_by(name, iso3) %>% 
  summarise(n_distinct(iso3), n()) %>%
  arrange(desc(n()))
```

(3). 
```{r}
min <- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarise(n = min(time))
```

```{r}
max <- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarise(n = max(time))
```

The minimum year in the data is `r min` and the maximum is `r max`. 

3. CO2 emossions 

(1)
```{r}
missing_co2 <- gapminder %>% 
  filter(is.na(co2)) %>% 
  group_by(time) %>% 
  summarise(n())
```

```{r}
missing_co2pc <- gapminder %>% 
  filter(is.na(co2_PC)) %>% 
  group_by(time) %>% 
  summarise(n())
```

1960 has the most missing data. 

(2). 
```{r}
gapminder %>% 
  filter(!is.na(co2)) %>% 
  filter(iso2 == "US" |
           name == "China" |
           name == "India" |
           name == "Japan") %>% 
  ggplot(aes(time, co2, col=factor(name))) +
  geom_line() +
  geom_point() +
  labs(x = "Time", y = "CO2", 
       col = "Country")
```

I can see that China has the most emssion during after 2000, and U.S has the relatively more emission
than those countries before that. 

(3)
```{r}
gapminder %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(iso2 == "US" |
           name == "China" |
           name == "India" |
           name == "Japan") %>% 
  ggplot(aes(time, co2_PC, col=factor(name))) +
  geom_line() +
  geom_point() +
  labs(x = "Time", y = "CO2_PC", 
       col = "Country")
```

There is a huge different between emission per capital than just emission. U.S now has 
the most ammount of emission per capital than other three countries. 

(4)
```{r}
gapminder %>% 
  filter(!is.na(co2_PC),
         time == 2016) %>% 
  group_by(region) %>% 
  summarize(mean(co2_PC))
```
I can see that Africa has the lowest average emission per capital, and Oceania
has the highest among the continents. 

(5) 
```{r}
gapminder %>% 
  filter(time %in% c(1960, 2016),
         !is.na(co2_PC),
         !is.na(region)) %>% 
  group_by(region, time) %>% 
  summarize(avg = mean(co2_PC)) %>% 
  ggplot(aes(region, avg, fill = as.factor(time))) +
  geom_col(position = "dodge")
```

6. 
```{r}
gapminder %>%
  filter(!is.na(co2_PC),
         !is.na(region)) %>% 
  group_by(region, name, .add = TRUE) %>%
  summarize(n()) %>% 
  group_split(region)
```

4. GDP per capital 
(1) 
```{r}
gapminder %>% 
  filter(time == 1960,
         !is.na(GDP_PC),
         !is.na(lifeExpectancy)) %>% 
  ggplot() +
  geom_point(aes(lifeExpectancy, GDP_PC, col = region, size = totalPopulation))
```
This scatterplot shows that European countries have relatively higher GDP per capital
and higher life expectancy. Asian countries have bigger size. 


(2)
```{r}
gapminder %>% 
  filter(time == 2019,
         !is.na(GDP_PC),
         !is.na(lifeExpectancy)) %>% 
  ggplot() +
  geom_point(aes(lifeExpectancy, GDP_PC, col = region, size = totalPopulation))
```

(3)
Comparing the two graphs, we can see that most of the countries increased their 
life expectancy, but the GDP per capital hasn't really differed from back then. 

(4)
```{r}
gapminder %>%
  filter(!is.na(lifeExpectancy),
         !is.na(region),
         time == 1960) %>% 
  group_by(region) %>%
  summarize(avg = mean(lifeExpectancy))
```

```{r}
gapminder %>%
  filter(!is.na(lifeExpectancy),
         !is.na(region),
         time == 2019) %>% 
  group_by(region) %>%
  summarize(avg = mean(lifeExpectancy))
```
I think the result pretty much matches the graphs that I computed above. 

(5)
```{r}
gapminder %>% 
  group_by(region, time) %>% 
  filter(!is.na(lifeExpectancy),
         !is.na(region)) %>% 
  summarize(avg = mean(lifeExpectancy)) %>% 
  arrange(start = 1960, end = 2019,
          rank(avg)) %>% 
  ungroup() %>% 
  mutate(growth = avg - lag(avg)) %>% 
  group_split(region)
```

I can see that for all of the continents, there is growth in life expectancy
within the year of 1960 to 2019. 

(6)
```{r}
gapminder %>% 
  filter(time %in% c(1960, 2016),
         !is.na(GDP_PC)) %>% 
  ggplot(aes(region, GDP_PC, fill = as.factor(time))) +
  geom_col(position = "dodge")
```

(7)
```{r}
gapminder %>%
  filter(time == 1960,
         !is.na(lifeExpectancy),
         !is.na(region)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  select(name, lifeExpectancy, rank) %>% 
  filter(name == "United States of America")
```




```{r}
gapminder %>%
  filter(time == 2019,
         !is.na(lifeExpectancy),
         !is.na(region)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  select(name, lifeExpectancy, rank) %>% 
  filter(name == "United States of America")
```

(8)
```{r}
gapminder %>%
  filter(time == 1960,
         !is.na(lifeExpectancy),
         !is.na(region)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  mutate(relative = rank /n_distinct(name)) %>% 
  select(name, lifeExpectancy, rank, relative) %>% 
  filter(name == "United States of America")
```

```{r}
gapminder %>%
  filter(time == 2019,
         !is.na(lifeExpectancy),
         !is.na(region)) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  mutate(relative = rank /n_distinct(name)) %>% 
  select(name, lifeExpectancy, rank, relative) %>% 
  filter(name == "United States of America")
```











